Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-24 paper.tex
Clarify matching vs unification

If $h : A$ is a hypothesis in the context of a goal and $A$ unifies with the type $B$ of slot $i$, we say that $h$ \emph{matches} $i$.
Unification induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $B[σ]$; we call this substitution $\sub_{i}(h)$.
In Sec.~\ref{sec:implementation}., we will specify exactly what \enquote{unifies} means.

@@

Note that matches represent only those partial applications where arguments are added without gaps.
Hence for a rule $r : A → B → C → D$, the applications $r~a~\_ : B \to D$ and $r~a~b : D$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to D$.
This will prevent us from constructing $r~a~b$ in two different but ultimately redundant ways.
The term orders used in various saturation calculi serve a similar function.

@@

Before this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
This approach treats forward rules essentially like any other Aesop rule.
Hence, when Aesop selects a forward rule $r$ with $m$ slots of types $A₁, \dots, Aₘ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.

To that end, we first unify the type of the last slot, $Aₘ$, with each hypothesis $h : B$ such that $B$ and $Aₘ$ likely unify (as determined by a discrimination tree index).
If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
We then unify $A_{m-1}[σ₁]$ with the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
If a premise $A_{m-i}[σᵢ]$ does not unify with any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
These may lead to a different substitution $τ$ and therefore to matching hypotheses for $A_{i}[τ]$.
For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis $h : B[σ]$, where $B$ is the conclusion of $r$.

This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
Otherwise, we perform a large number of unifications between premises and hypotheses (though some of these could perhaps be avoided heuristically~\cite{TammetSubsumption}).
Furthermore, the whole process is repeated when we proceed to another goal with a similar context.
In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be substantially similar, if not identical, to the current context, but these similarities are not exploited by the naive algorithm.

@@

\para{Unification.}
In Lean's type theory, types contain programs and unification is expected to respect definitional equality, i.e.\ equality up to evaluation.
For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
As a result, unifying premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
Our technique therefore seeks to minimise the number of unifications performed.

To limit the expense of unification, we perform unification at \emph{reducible} transparency.
Lean has multiple transparency levels that determine which constants are unfolded.
\enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions.
Additionally, we use Lean's usual approximation of higher-order unification, which is complete for first-order unification problems and heuristically solves some higher-order problems.
Either of these limitations render forward reasoning incomplete, but since the same unification is used throughout Lean, users are used to this.

During unification, we treat metavariables in hypotheses as constants, i.e.\ we never assign them.
This reflects the expected behaviour of forward rules: given a hypothesis $h : \mvar{x} < \mvar{y}$ and rule $1 < 0 → ⊥$, applying the rule to $h$ would replace $\mvar{x}$ with 1 and $\mvar{y}$ with 0 in all goals where $\mvar{x}$ and $\mvar{y}$ appear, even though this may render some goals unprovable.

@@

There are reasons to believe that this scheme is more efficient than alternative schemes relying on repeated unifications.
For example, we could use Lean's discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
However, these discrimination trees are imperfect, so the validity of a lookup result must be verified by unifying the query expression with the expression that was originally stored in the discrimination tree.
We believe that performing these unifications would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).

@@

Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals for the benchmarks.
Unifying two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
We can therefore use moderately large numbers to simulate the bigger unification problems that Aesop encounters, for example, in Mathlib's category theory development.

>>>>

If $h : C$ is a hypothesis and $C$ matches the type $A$ of slot $i$, we say that $h$ \emph{matches} $i$.
Matching induces a substitution $σ$ such that $A[σ]$ is definitionally equal to $C$; we define $\sub_{i}(h) ≔ σ$.
What exactly \enquote{matching} means is discussed in Sec.~\ref{sec:implementation}.

@@

Note that matches represent only those partial applications where arguments are added without gaps.
Hence for a rule $r : A → B → C → D$, the applications $r~a~\_ : B \to D$ and $r~a~b : D$ for some $a : A$ and $b : B$ correspond to matches, but not $r~\_~b : A \to D$.
This prevents us from constructing $r~a~b$ in two different but ultimately redundant ways.
The term orders used in various saturation calculi serve a similar function.

@@

Before this work, Aesop used what we call \emph{naive forward reasoning}~\cite{Aesop}.
This approach treats forward rules essentially like any other Aesop rule.
Hence, when Aesop selects a forward rule $r$ with $n$ slots of types $A₁, \dots, Aₙ$ as the highest-priority rule currently applicable to a goal $Γ ⊢ T$, we look for all complete matches for $r$ in $Γ$.

To that end, we first match the type of the last slot, $Aₙ$, with each hypothesis $h : B$ (using a discrimination tree index to quickly determine potentially matching hypotheses).
If successful, this results in a substitution $σ₁$ containing all variables that occur in $Aₘ$.
We then match $A_{m-1}[σ₁]$ against the type of each hypothesis in the context, yielding a substitution $σ₂$, and proceed with $A_{m-2}[σ₂]$, etc., until we obtain a complete match for $r$.
If a premise $A_{m-i}[σᵢ]$ does not match any hypothesis type, we backtrack and consider other potential matching hypotheses for $A_{i+1}$.
These may lead to a different substitution $τ$ and therefore to different matching hypotheses for $A_{i}[τ]$.
For each complete match $m$ with substitution $σ$ found during this process, we add a new hypothesis of type $B[σ]$, where $B$ is the conclusion of $r$.

This approach is efficient when the last slot of a rule is not present in the context since the rule is then excluded by indexing and never run.
Otherwise, we match large numbers of premises and hypotheses against each other (though this could likely be sped up heuristically~\cite{TammetSubsumption}).
Moreover, the whole process is repeated when we proceed to another goal with a similar context.
In particular, the contexts of subgoals produced by whatever rule is eventually applied to the current goal are likely to be similar to the current context, but the naive algorithm does not exploit these similarities.

@@

\para{Matching.}
In type theory, types contain programs and matching (or, more generally, unification) is expected to respect definitional equality, i.e.\ equality up to evaluation.
For instance, a theorem with premise $P~2$, where $P$ is some predicate, should apply to a hypothesis of type $P~((\Lam{x}{x})~2)$ since the two types are definitionally equal.
As a result, matching premises and hypotheses is arbitrarily expensive in theory and a major cost centre in practice.
We therefore seek to avoid redundant invocations of the matching procedure.

To limit the cost of matching, we match with \emph{reducible transparency}.
Lean has multiple transparency levels that determine which constants are unfolded.
\enquote{Reducible} is the most restrictive of these, so we unfold few (and, in practice, only non-recursive) definitions.
Additionally, we use Lean's usual approximation of higher-order matching, which is complete for first-order matching problems and heuristically solves some higher-order problems.
Either of these limitations render forward reasoning incomplete, but since the same matching algorithm is used throughout Lean, users are used to this source of incompleteness.

@@

There are reasons to believe that this scheme is more efficient than alternative schemes relying on repeated unifications.
For example, we could use Lean's discrimination trees, which implement a map from expressions to arbitrary data where lookup respects definitional equality.
However, these discrimination trees are imperfect, so the validity of a lookup result must be verified by matching the query expression against the expression that was originally stored in the discrimination tree.
We believe that this would generally be more expensive than calculating the RPINF of each hypothesis type (which can be cached).

@@

Instead of the natural number type of Lean's standard library, which has special support in the compiler and kernel, we use a custom type of Peano naturals for the benchmarks.
Matching two concrete Peano natural numbers $n$ and $m$ involves comparing $\min(n, m)$ constructors.
We can therefore use moderately large numbers to simulate the bigger matching problems that Aesop encounters, for example, in Mathlib's category theory development.
